Continuous-Time Markov Chain with Alarms analysis and comparison of tools \thanks{This paper has been written in the context of my internship at the Faculty of Informatics with Vojtech Reh{\'{a}}k at the Masaryk University}

### 2018-07-24 Report/report/paper.tex
last modif on institution paper

\begin{document}
%
\title{Continuous-Time Markov Chain with Alarms analysis and comparison of tools \thanks{Thanks t}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Thomas Mari\inst{1,2}}%\orcidID{0000-1111-2222-3333}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{ENS-Rennes, Campus de Ker lann, Avenue Robert Schuman, 35170 Bruz, France \and
Masaryk University, Botanická 68A, 602 00 Brno-Královo Pole, Czech republic
\email{thomas.mari@ens-rennes.fr}\\
\url{http://perso.eleves.ens-rennes.fr/people/thomas.mari/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
For the analysis of ACTMC, Phase-type fitting is the advised way of modeling non-Markovian distributions within PRISM model checker. We ran experimental computations and deduced a reliable way of obtaining precise analysis results for phase-type fitted PRISM CTMCs.
We aim to compare Phase-type fitting with other approaches that handle directly the non-Marlovian distributions in term of efficiency and precision.

@@

	\begin{figure}
		\begin{center}
			\includegraphics[trim=0.2cm 0cm 0cm 1.35cm,width=10cm, clip]{../picture/New_model/1E-6/performance_dynamic.png}
		\end{center}
		\caption{Time of computation in seconds of the steady-state probabilities against various values of $k$ for lowered and adjusted absolute termination epsilon $\epsilon = 10^{-10}$, i.e. corresponding to the data shown in Figure \ref{fig:distance_dynamic_1E-10_constant_kappa}. The computation of our PRISM ACTMC is constant in regards to $k$ because it does not use phase-type fitting. }
		\label{fig:performance_dynamic}
	\end{figure}

@@

	For better performance, we suggest usage of more specialized tools that can deal with non-Markovian distributions directly, without phase-type fitting. Our PRISM ACTMC extension \footnote{Our PRISM ACTMC extension is not fully completed yet and lacks optimization. It will be officially released later, but early development version is available at \href{https://github.com/VojtechRehak/prism-gsmp}{https://github.com/VojtechRehak/prism-gsmp}.} has delivered results with sufficient precision a lot faster.

	This experimental statement strengthens the conjecture that further heuristics can increase the efficiency of an analysis with a non-Markovian distributions\cite{DBLP:journals/corr/KorenciakKR14}. To some extent GSMP-PRISM is challenging the PH fitting approach. Hence it could be interesting to compare the efficiency of GSMP-PRISM with the analysis of SPN with the tool Oris.

>>>>

\begin{document}
%
\title{Continuous-Time Markov Chain with Alarms analysis and comparison of tools \thanks{This paper has been written in the context of my internship at the Faculty of Informatics with Vojtech Reh{\'{a}}k at the Masaryk University}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Thomas Mari\inst{1,2}}%\orcidID{0000-1111-2222-3333}}
%
\titlerunning{ACTMC analysis and comparison of tools}% Part of RIGHT running header
\authorrunning{Thomas Mari}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{ENS-Rennes, Campus de Ker lann, Avenue Robert Schuman, 35170 Bruz, France \and
Masaryk University, Botanická 68A, 602 00 Brno-Královo Pole, Czech republic
\email{thomas.mari@ens-rennes.fr}\\
\url{http://perso.eleves.ens-rennes.fr/people/thomas.mari/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
For the analysis of ACTMC, Phase-type fitting is the advised way of modeling non-Markovian distributions within PRISM model checker. We ran experimental computations and deduced a reliable way of obtaining precise analysis results for phase-type fitted PRISM CTMCs.
We aim to compare Phase-type fitting with other approaches that handle directly the non-Marlovian distributions in term of efficiency and precision.

@@

	\begin{figure}
		\begin{center}
			\includegraphics[trim=0.2cm 0cm 0cm 1.35cm,width=10cm, clip]{../picture/New_model/1E-6/performance_dynamic.png}
		\end{center}
		\caption{Time of computation in seconds of the steady-state probabilities against various values of $k$ for lowered and adjusted absolute termination epsilon $\epsilon = 10^{-10}$, i.e. corresponding to the data shown in Figure \ref{fig:distance_dynamic_1E-10_constant_kappa}. The computation of our PRISM ACTMC is constant in regard to $k$ because it does not use phase-type fitting. }
		\label{fig:performance_dynamic}
	\end{figure}

@@

	For better performance, we suggest usage of more specialized tools that can deal with non-Markovian distributions directly, without phase-type fitting. Our GSMP PRISM extension \footnote{Our GSMP PRISM extension is not fully completed yet and lacks optimization. It will be officially released later, but early development version is available at \href{https://github.com/VojtechRehak/prism-gsmp}{https://github.com/VojtechRehak/prism-gsmp}.} has delivered results with sufficient precision a lot faster.

	This experimental statement strengthens the conjecture that further heuristics can increase the efficiency of an analysis with a non-Markovian distributions\cite{DBLP:journals/corr/KorenciakKR14}. To some extent GSMP-PRISM is challenging the PH fitting approach. Hence, it could be interesting to compare the efficiency of GSMP-PRISM with the analysis of SPN with the tool Oris.

### 2018-07-24 Report/report/preliminaries.tex
last modif on institution paper

We will go through the theoretical bases needed to understand the contribution. For a complete understanding you can refer to the chapter two of this thesis \cite{Korenciak2018thesis}.
\subsection{Time distribution}
The class of models we use is probabilistic continuous time model, which means that the time passed in a state will depend on a time distribution. A time distribution can be depicted by its cumulative distribution function (CDF) $F : \mathbb{R} \rightarrow[0,1]$, $F(t)$ is the probability that the event will happend before the time $t$. The exponential distribution of rate $\lambda$ has a CDF $F(t) = 1 - \exp(-\lambda t)$, it is important to keep in mind that a rate refers to the exponential distribution. The Dirac distribution of timeout $\tau$ has a CDF $F(t) = 
\left\{
	\begin{array}{l}
		0 \text{ if } t < \tau\\
		1\text{ else}
	\end{array}
\right.$. The keyword deterministic is used when referring to so event following a Dirac distribution.
The Erlang distribution can be seen as the distribution of a sequence of $k$ exponential distribution of parameter $\lambda$. Hence, the Erlang distribution has two parameter $k$ and $\lambda$. 
In this paper, time distribution which are not exponential will be called non-Markovian distribution. Hence, a non-Markovian model will refer to a model with non-Markovian distribution.
\subsection{CTMC}
\label{CTMC_def}
A continuous-time Markov chain (CTMC) is a triple $C = (S,Q,s_{in})$, where 
\begin{itemize}
	\item[$\bullet$] S is a finite set of states
	\item[$\bullet$] $Q : S \times S \rightarrow \mathbb{R}_{\geq 0}$ is a matrix of rates such that $\sum_{s' \in S} Q(s,s')  > 0$ for each $s \in S$
	\item[$\bullet$] $s_{in} \in S$ is an initial state
\end{itemize} 
$Q(s,s')$ is the rate of the transition from $s$ to $s'$.
We define the exit rate of state $s$ as $\lambda_s = \sum_{s' \in S} Q(s,s')$

@@

%Continuous-Time Markov Chain with Alarms (ACTMC)s. I won't give the formal semantics which might not fit in this paper. 
We can see ACTMC as CTMC with some events called alarms that are enabled in disjoints sets of states and those events has their own timer which is active only when inside the enabled set and happend according to some continuous time distribution. Those alarms works concurrently with the exponential transition seen in the CTMC.

@@

%CASE 2
If $\eta_i > t_i$, the alarm does not ring. 
$t_i$ the value of delay remain the same and match the time spent in $s_i$.
$s_{i+1}$ is chosen according to the discrete distribution $\dfrac{Q(s_i,.)}{\lambda_{s_i}}$.
If $s_{i+1} \in S_{off}$, then  $\eta_{i+1} = \infty$.
If $s_{i+1} \in S_{a}$ and $s_{i} \in S_{a}$, then  $\eta_{i+1} =\eta_i - t$.
If $s_{i+1} \in S_{a}$ and $s_{i} \notin S_{a}$, then
$\eta_{i+1}$ is randomly chosen according to $F_a$. 
If the state remain in the set enabled by the alarm, the timer is updated. Otherwise it is reset according to the new enabled set.

>>>>

We will go through the theoretical bases needed to understand the contribution. For a complete understanding you can refer to the chapter two of this thesis \cite{Korenciak2018thesis}.
\subsection{Time distribution}
The class of models we use is probabilistic continuous time model, which means that the time passed in a state will depend on a time distribution. A time distribution can be depicted by its cumulative distribution function (CDF) $F : \mathbb{R} \rightarrow[0,1]$, $F(t)$ is the probability that the event will happen before the time $t$. The exponential distribution of rate $\lambda$ has a CDF $F(t) = 1 - \exp(-\lambda t)$, it is important to keep in mind that a rate refers to the exponential distribution. The Dirac distribution of timeout $\tau$ has a CDF $F(t) = 
\left\{
	\begin{array}{l}
		0 \text{ if } t < \tau\\
		1\text{ else}
	\end{array}
\right.$. The keyword deterministic is used when referring to so event following a Dirac distribution.
The Erlang distribution can be seen as the distribution of a sequence of $k$ exponential distribution of parameter $\lambda$. Hence, the Erlang distribution has two parameter $k$ and $\lambda$. 
In this paper, time distribution which are not exponential will be called non-Markovian distribution. Hence, a non-Markovian model will refer to a model with non-Markovian distribution.
\subsection{CTMC}
\label{CTMC_def}
A continuous-time Markov chain (CTMC) is a triple $C = (S,Q,s_{in})$, where 
\begin{itemize}
	\item[$\bullet$] S is a finite set of states
	\item[$\bullet$] $Q : S \times S \rightarrow \mathbb{R}_{\geq 0}$ is a matrix of rates such that $\sum_{s' \in S} Q(s,s')  > 0$ for each $s \in S$
	\item[$\bullet$] $s_{in} \in S$ is an initial state
\end{itemize} 
$Q(s,s')$ is the rate of the transition from $s$ to $s'$.
We define the exit rate of state $s$ as $\lambda_s = \sum_{s' \in S} Q(s,s')$

@@

%Continuous-Time Markov Chain with Alarms (ACTMC)s. I won't give the formal semantics which might not fit in this paper. 
We can see ACTMC as CTMC with some events called alarms that are enabled in disjoints sets of states and those events has their own timer which is active only when inside the enabled set and happend according to some continuous time distribution. Those alarms work concurrently with the exponential transition seen in the CTMC.

@@

%CASE 2
If $\eta_i > t_i$, the alarm does not ring. 
$t_i$ the value of delay remain the same and match the time spent in $s_i$.
$s_{i+1}$ is chosen according to the discrete distribution $\dfrac{Q(s_i,.)}{\lambda_{s_i}}$.
If $s_{i+1} \in S_{off}$, then  $\eta_{i+1} = \infty$.
If $s_{i+1} \in S_{a}$ and $s_{i} \in S_{a}$, then  $\eta_{i+1} =\eta_i - t$.
If $s_{i+1} \in S_{a}$ and $s_{i} \notin S_{a}$, then
$\eta_{i+1}$ is randomly chosen according to $F_a$. 
If the state remain in the set enabled by the alarm, the timer is updated. Otherwise, it is reset according to the new enabled set.

### 2018-07-24 Report/report/related_work.tex
last modif on institution paper

A less expressive model is the CTMC defined in Subsection \ref{CTMC_def}, PRISM\cite{KNP11} and Storm \cite{DBLP:journals/corr/DehnertJK017} are both well known tools to analyze CTMC. The engine of those tools are quite efficient for analysis due to the memoryless property of the exponential distribution. Phase-type (PH) approximation can be used to approach a non-Markovian model. The downside of PH approximation is the increase of the number of states which can slow down the process of analysis.  PH Fitting can be challenging depending on the PH distribution\cite{DBLP:journals/corr/KorenciakKR14}.
There is kinds of distributions which requires a lot of phases to be accurately approached by PH fitting. One of those distribution is a shifted exponential distribution, which can be seen as a sequence of a deterministic timeout and a exponential distribution. The motivation of using those distribution is to depict better some events like transmissions that can't happened immediately due to physical limits.

>>>>

A less expressive model is the CTMC defined in Subsection \ref{CTMC_def}, PRISM\cite{KNP11} and Storm \cite{DBLP:journals/corr/DehnertJK017} are both well known tools to analyze CTMC. The engine of those tools are quite efficient for analysis due to the memoryless property of the exponential distribution. Phase-type (PH) approximation can be used to approach a non-Markovian model. The downside of PH approximation is the increase of the number of states which can slow down the process of analysis.  PH Fitting can be challenging depending on the PH distribution\cite{DBLP:journals/corr/KorenciakKR14}.
There are kinds of distributions which require a lot of phases to be accurately approached by PH fitting. One of those distributions is a shifted exponential distribution, which can be seen as a sequence of a deterministic timeout and a exponential distribution. The motivation of using those distributions is to depict better some events like transmissions that can't happen immediately due to physical limits.
